---
title: "Teaching R in DASC 1104"
subtitle: "R Part I"
author: "John Tipton"
institute: "The University of Arkansas"
date: "2021/06/07 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      navigation:
        scroll: false      
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
urlcolor: blue
knit: pagedown::chrome_print
---

```{css, echo = FALSE}
.remark-slide-content {
  font-size: 18px;
  padding: 20px 80px 20px 80px;
}
.remark-code, .remark-inline-code {
  background: #f0f0f0;
}
.remark-code {
  font-size: 14px;
}
.huge .remark-code { /*Change made here*/
  font-size: 200% !important;
}
.very-large .remark-code { /*Change made here*/
  font-size: 150% !important;
}
.large .remark-code { /*Change made here*/
  font-size: 125% !important;
}
.small .remark-code { /*Change made here*/
  font-size: 75% !important;
}

.very-small .remark-code { /*Change made here*/
  font-size: 50% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 40% !important;
}
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Materials

- Available on gitHub at [https://github.com/jtipton25/dasc-1104-teaching](https://github.com/jtipton25/dasc-1104-teaching) 

---

# Guiding principles

* Make the initial burden to working with real data as minimal as possible

--

* Teach a **process** of working with data more than the actual programming

    * Focus on actions/principles rather than specific syntax
    
    * More than in other programming classes, focus on [reproducibility](https://rstats.wtf/save-source.html#use-an-ide)
    

* Great resources for learning R/python/git/shell

    * [https://software-carpentry.org/lessons/index.html](https://software-carpentry.org/lessons/index.html)

--


* Scaffold Learning objectives and provide many examples
    
    * Here is a dataset -- Generate a scatterplot of mpg vs. engine displacement (demo with `mtcars`)
    
    * Start with simple examples and show how to build
    
    * Encourage student participation by working examples alongside 

---

# Guiding principles

* Allow for student creativity and spontaneity

    * Different from typical programming classes
    
    * Here is a dataset -- Create two visualizations that tell me something interesting about it
    
--

* Focus on the high-level concepts before getting into the details

    * Some low level knowledge is ok -- variable types
    
    * Other low level knowledge not needed at this stage
    
        * Parsing files one line at a time, etc.
        
        


---


# The Rstudio IDE

- I recommend the [RStudio IDE](https://www.rstudio.com/products/rstudio/), but you do not have to.

    - Allows for R and python code development
    
* Can be installed [locally](https://www.rstudio.com/products/rstudio/) or on a [server](https://www.rstudio.com/products/workbench/) (I use a SSO server option)

* [RStudio Cloud](https://rstudio.cloud/) provides a remote-hosting option for teaching

---

# Project-oriented workflows

- Especially useful for students with less comp-sci experience
    
- Makes file path management easier
    
- References [here](https://rstats.wtf/project-oriented-workflow.html) and [here](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) and [here](https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/)



## What does this code do?

```{r, eval = FALSE}
library(tidyverse)
setwd("~/Documents/John/hw-1")
dat <- read_csv("my-secrets.csv")
```

--

* What might be the issue with this code?

---

# Hands-on experience project management in Rstudio

* Use of Rstudio projects (5-10 minutes)

* Use of the `here` library and the `here()` function

* Namespaces in `R`

* Generate "data file" in excel, load and process the data

* For more details, see Lecture 9 in unit 4




---

# The Two Towers

- The R language has split into two dialects "base R" and the "tidyverse"

- [Much](http://varianceexplained.org/r/teach-tidyverse/) has [been written](https://blog.ephorie.de/why-i-dont-use-the-tidyverse) about this difference

- Having learned Base R first and `tidyverse` second -- I am a strong proponent of the `tidyverse` -- there is a strong movement to [translate this syntax to python](https://srobert4.github.io/post/2019-08-08-tidyversetopandas/)

    - Most of my students would agree that `tidyverse` is the way to go
    
    - [Many python users](https://stmorse.github.io/journal/tidyverse-style-pandas.html) also agree
    
    
---

# The Rings of Power


```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
glimpse(penguins)
```


---

# The Rings of Power

* Create a histogram of `body_mass_g`

.pull-left[
* Using base R


```{r}
hist(penguins$body_mass_g)
```
]
.pull-right[
* Using `tidyverse`

```{r, warning = FALSE, message = FALSE}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram()
```
]

---

# But they were all of them decieved

* Create a histogram of `body_mass_g` for each `species` and `sex`

.pull-left[
* Using base R
.very-small[

```{r}
layout(matrix(1:6, 2, 3))
hist(penguins[penguins$sex == "male" & penguins$species == "Adelie", ]$body_mass_g)
hist(penguins[penguins$sex == "male" & penguins$species == "Chinstrap", ]$body_mass_g)
hist(penguins[penguins$sex == "male" & penguins$species == "Gentoo", ]$body_mass_g)
hist(penguins[penguins$sex == "female" & penguins$species == "Adelie", ]$body_mass_g)
hist(penguins[penguins$sex == "female" & penguins$species == "Chinstrap", ]$body_mass_g)
hist(penguins[penguins$sex == "female" & penguins$species == "Gentoo", ]$body_mass_g)
```
]
]
.pull-right[
* Using `tidyverse`

```{r, warning = FALSE, message = FALSE}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram() +
  facet_grid(sex ~ species)
```
]

---

# For in secret the Dark Lord Sauron forged in secret a Master Ring, to control all others. 

* How many penguins are in the dataset of each `sex` and `species`?

.pull-left[
```{r}
with(penguins, table(sex, species))
```
]

.pull-right[
```{r}
penguins %>%
  count(sex, species)
```
]

* Where did the `NA` values go?


---

# One ring to rule them all!

* What does this code do?

.small[
```{r, eval = FALSE}
aggregate(airquality[, "Ozone"], 
          list(Month = airquality[, "Month"]), 
          mean, na.rm = TRUE)
```
]
--
.small[
```{r, echo = FALSE}
aggregate(airquality[, "Ozone"], 
          list(Month = airquality[, "Month"]), 
          mean, na.rm = TRUE)
```
]
--
.small[
```{r, eval = FALSE}
airquality %>% 
  group_by(Month) %>% 
	summarize(mean_o3 = mean(Ozone, na.rm = TRUE))
```
]
--
.small[
```{r, echo = FALSE}
airquality %>% 
  group_by(Month) %>% 
	summarize(mean_o3 = mean(Ozone, na.rm = TRUE))
```
]

--

* Which example is more **human readable?** 

---

# Working with RMarkdown

* Demo with RMarkdown examples (15-20 minutes)

.pull-left[
* Code chunks

* Running code (ctrl-enter, run all above)

* installing vs. loading packages

* Restarting the R environment and clearing variables

    * Do this often

* **Settings**--don't save your workspace

* Plain text, math equations, and communication

* Code chunk options


]

.pull-right[

* File paths and loading data

* Managing the project environment

* Formatting and troubleshooting

* The tinytex/latex problem

    * `install.packages('tinytex')
    * `tinytex::install_tinytex()`

* python code in RMarkdown

* sourcing in scripts

* live preview (I'm learning this in real time)

* Getting help with `?`
]


---

# What to teach

- Programming concepts are important

    - loops, data structures, functions
    
- But, I **don't** teach R as if it were a programming language
    - R is a programming language for **data science**

    - Many "real" programmers don't like R because it is not a "progamming first" language
    
        - R is a **data-science** first language, don't get mad when it isn't what you want it to be
    
    - Show how to do data science using R as a tool
    
    
- Lead with data visualization, follow up with data manipulation

--- 

# Explore for meaningful datasets to use

    
1) [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday)
    
2) [FiveThirtyEight](https://github.com/fivethirtyeight/data)

3) [BuzzFeed News](https://github.com/BuzzFeedNews/everything)
    
4) [Kaggle](https://www.kaggle.com/datasets)

5) [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets.html)

---

# Focus on basic syntax and datatypes (demo)

## Typically in class, I include code in the slides, but not here

- vectors

- matrices

- arrays

- lists

- factors

- data.frames

---


# Atomic types (demo)

- logical/boolean

- numeric

- integer

- complex (not covered)

- character/string


---

# Working with data

* what are data.frames?

* accessing rows/columns with `[`

* accessing variables with `[`

* accessing variables with `$`

---












