---
title: "Teaching R in DASC 1104"
subtitle: "R Part II"
author: "John Tipton"
institute: "The University of Arkansas"
date: "2021/06/07 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      navigation:
        scroll: false      
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
urlcolor: blue
knit: pagedown::chrome_print
---

```{css, echo = FALSE}
.remark-slide-content {
  font-size: 18px;
  padding: 20px 80px 20px 80px;
}
.remark-code, .remark-inline-code {
  background: #f0f0f0;
}
.remark-code {
  font-size: 14px;
}
.huge .remark-code { /*Change made here*/
  font-size: 200% !important;
}
.very-large .remark-code { /*Change made here*/
  font-size: 150% !important;
}
.large .remark-code { /*Change made here*/
  font-size: 125% !important;
}
.small .remark-code { /*Change made here*/
  font-size: 75% !important;
}

.very-small .remark-code { /*Change made here*/
  font-size: 50% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 40% !important;
}
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


# Learning R

- [Introduction to tidy statistics](https://www.rstudio.com/resources/webinars/a-gentle-introduction-to-tidy-statistics-in-r/)

- [RStudio primers](https://rstudio.cloud/learn/primers)

- [Software carpentries](https://software-carpentry.org/lessons/index.html)

    - [R for Reproducible Analyses](https://swcarpentry.github.io/r-novice-gapminder/)
    
    - [Programming with R](https://swcarpentry.github.io/r-novice-inflammation/)

- [R for data science](https://r4ds.had.co.nz/)


---


# Materials 

- Available on gitHub at [https://github.com/jtipton25/dasc-1104-teaching](https://github.com/jtipton25/dasc-1104-teaching)

---

# Course make file

* In the loaded `dasc1104-teaching` project, the following command will generate all code


```{r, eval = FALSE}
# note: this will take a while, maybe 15-20 minutes
source(here::here("make.R"))
```

# Lectures

- unit_4_09_intro_R_Rmarkdown.Rmd

- unit_4_R_10_intro_plotting.Rmd

- unit_4_R_11_File_IO_and_basic_data_wrangling.Rmd

- unit_4_R_12_Data_types_and_objects.Rmd

- unit_4_R_13_Functional_programming.Rmd

- unit_4_R_14_Intro_to_modeling.Rmd

---

## Brief recap of each unit

---

# Intro to R and Rmarkdown

- Focus on reproducibility and workflow

- Heavily inspired by [R for Reproducible Analyses](https://swcarpentry.github.io/r-novice-gapminder/)

- Much of this was covered in part I

---

# Data visualization and plotting

- Goal: to get students interested and excited when working with data

    - Give enough background and language for students to seek out help online for data visualization
    
    - Emphasis on flexibility and exploration of data through visualization
    
    - Worked through a couple datasets to demonstrate how the visualization process works


---

# Data Visualization

- Using `ggplot` and the `mtcars` data, generate a scatterplot of fuel milage (`mpg`) as a function of engine power (`hp`)

- Explore how other variables in the `mtcars` dataset influence fuel milage

---

# File IO

- Read the `game_goals.csv` file from the `data` folder with `read_csv()` (from the `tidyverse` library) and give this `data.frame` the name `dat`

- Make sure you properly define the filepath

- Many more specifics about reading files
    - skipping lines
    - defining column variable types
    - different file delimiters
    - missing values and `NA`s
    - data comments

---

# Data manipulation

## Most important `dplyr` (a part of the `tidyverse`) commands

* Using pipes `%>%`

* Choose observations (rows) based on conditional values with `filter()`

* Reorder the variables with `arrange()`

* Select variables by name with `select()`

* Create new variables with `mutate()`

* Create summary variables with `summarize()` 
    * `summarise()` if you are British 
    
* Group variables with `group_by()`    
    

# Pipes

> The itsy bitsy spider went up the water spout

> Down came the rain and washed the spider out

> Out came the sun and dried up all the rain

> And the itsy bitsy spider went up the spout again

* multiple variable names, likely to make a typo

```
itsy_bitsy_spider_1 <- went(itsy_bitsy_spider, up = "waterspout")
itsy_bitsy_spider_2 <- rain_down(itsy_bitsy_spider_1, washed = "out")
itsy_bitsy_spider_3 <- sun(itsy_bitsy_spider_2, dried = "rain")
itsy_bitsy_spider_4 <- went(itsy_bitsy_spider_3, up = "spout again")
```


* overwriting the data, likely to make errors

```
itsy_bitsy_spider <- went(itsy_bitsy_spider, up = "waterspout")
itsy_bitsy_spider <- rain_down(itsy_bitsy_spider, washed = "out")
itsy_bitsy_spider <- sun(itsy_bitsy_spider, dried = "rain")
itsy_bitsy_spider <- went(itsy_bitsy_spider, up = "spout again")
```

---

# Pipes

> The itsy bitsy spider went up the water spout

> Down came the rain and washed the spider out

> Out came the sun and dried up all the rain

> And the itsy bitsy spider went up the spout again

* hard to read

```
went(
    sun(
        rain_down(
            went(itsy_bitsy_spider, up = "waterspout"),
            washed = "out"),
        dried = "rain"),
    up = "spout again"
)
```

* Much better

```
itsy_bitsy_spider %>%
    went(up = "waterspout") %>%
    rain_down(washed = "out) %>%
    sun(dried = "rain") %>%
    went(up = "spout again")
```

---

# Pipes

* How does the pipe work?

```{r}
x <- 1:10
x %>%
  mean()
```

is equivalent to 

```{r}
x %>%
  mean(.)
```

where the `.` is a placeholder for the variable `x` which results in 

```{r}
mean(x)
```



---

# Pipes

* Pipes are not always the optimal choice

* Try to avoid super long chains of pipes

    * Instead: break up the pipe into intermediate objects

* Pipes can't handle multiple inputs/outputs

--

* How to build a chain of pipes
    
    * Start by outlining a plan of attack
    
    * Write one small subset of the pipe chain at a time and verify the results


* Challenge: Ask me questions about the `diamonds` data


---

# Data manipulation

- Using the NHL goals dataset from before (named `dat`), `filter()` out all records where the `player` is `"Alex Ovechkin"` or `"Sidney Crosby"`. How many observations are there?

- Using the NHL goals dataset, `arrange()` the observations in decreasing order (the `desc()` function) using the `penalty_min` variable. Who on the list has two games in the top 10 penalty minutes of the dataset?

- Using the NHL dataset, `select()` the variables `player`, `shots`, and `goals`. Then, `mutate()` a new variable `shots_per_goal` by dividing the number of `shots` by the number of `goals`. Did you get any warnings/errors?

- Using the NHL dataset, `summarize()` the total `goals` for each `player`. Using `arrange()`, who are the top 5 players in all time goals (in the data)?


---

# Relational data

- Joining multiple datasets together

- `left_join()`

- `right_join()`

- `inner_join()`

- `outer_join()`


---

# Processing strings

- Use the `stringr` package (part of the `tidyverse`)

- Can use regular expressions

```{r}
(string1 <- "This is a string")
(string2 <- 'This is also a string')
(string3 <- 'This is a "quoted" string')
``` 

---

# String processing

- `str_length()`

- `str_c()`

- `str_sub()`

- others

---

# Factors

- Factors are an atomic type for categorical data

    - By default, these are unordered
    
    - Ordered factors can be used (good - better - best)
    
- `factor()`

- `levels()`

- `fct_recode()`

- `fct_order()`


---


# Dates (not covered)


---

# Functional programming (currently not covered, but should be)

- Code is for **humans** not computers

- Functions make code readable to humans


---

# Keep it DRY


* Don't Repeat Yourself

* If you copy and paste code 3 or more times, write a function

* Generate some data 

```{r}
df <- data.frame(
  x = rnorm(10), 
  y = rnorm(10),
  z = rnorm(10)
)
```

* The above isn't DRY, but there is no chance of errors and the code is easy to understand

---

# DRY

* Transform each variable to mean 0, standard deviation 1

```{r}
x_scaled <- (df$x - mean(df$x, na.rm = TRUE)) / sd(df$x, na.rm = TRUE)
y_scaled <- (df$y - mean(df$y, na.rm = TRUE)) / sd(df$y, na.rm = TRUE)
z_scaled <- (df$z - mean(df$z, na.rm = TRUE)) / sd(df$y, na.rm = TRUE)
```
    
--

* Can you spot the error?

--

```{r, eval = FALSE}
z_scaled <- (df$z - mean(df$z, na.rm = TRUE)) / sd(`df$y`, na.rm = TRUE)
```


* write a function to do this

```{r}
rescale_variable <- function(x) {
  mean_x <- mean(x, na.rm = TRUE)
  sd_x   <- sd(x, na.rm = TRUE) 
  ## save the variable for the output
  out <- (x - mean_x) / sd_x
  return(out)
}
```

* Much less likely to make an error

```{r}
x_scaled <- rescale_variable(df$x)
y_scaled <- rescale_variable(df$x)
z_scaled <- rescale_variable(df$z)
```

    
---

# Naming Functions

* Name functions for human readers!

```{r, eval = FALSE}
# bad name -- what does this do?
f()

# what does this do?
func()

# longer names -- use autocomplete
fill_missing_values()
pre_process_data()
```

* Don't name your functions over functions or variables which already exist!

```{r, eval = FALSE}
mean <- function(x) {}
TRUE <- function(x, y) {}
```

---

# Naming Functions

* Pick a naming convention and stick with it (I prefer snake case)

```{r}
# snake case
my_function_name <- function(x, y) {}
# camel case
myFunctionName <- function(x, y) {}
```

* Try to reuse leading names -- allows for autocomplete

```{r, eval = FALSE}
# better
check_input()
check_parameters()
check_inits()

# not as good
input_check()
parameters_check()
inits_check()
```

---    

# For loops

- for loops in R

```{r}
value <- 0
for (i in 1:100) {
  value <- value + i
}
value
```

---

# Writing functions

- given a `data.frame`, calculate the mean of each column

- write a function that returns the mean of each column


---

# The project

- Idea: have each student construct a blog and perform a basic analysis using the skills learned in the class

- Maybe next semester the project will be smaller in scope to cover more material

- Have students use data to tell a story using summary statistics and data visualization

---

# Example projects


- https://jdkincan.netlify.app/about/
- https://arbeger-blog.netlify.app/
- https://dasc-1104-alc.netlify.app/
- https://dasc1104-blog-project.netlify.app/
- https://tracef.netlify.app/
- https://abbyharris.netlify.app/
- https://akoster-dasc-1104-blog.netlify.app/
- https://dasc-1104-bmarlow.netlify.app/
- https://hpmcdona.netlify.app/
- https://mrrice-dasc-1104-blog.netlify.app/
- https://ntrogers.netlify.app/about/
- https://nifty-wozniak-e3e1d9.netlify.app/about/
- https://dasc-1104-calebwilliamsblog.netlify.app/about/
- https://rjyates-dasc-blog.netlify.app/


