---
title: "Data Visualization with ggplot2"
author: "John Tipton"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      navigation:
        scroll: false      
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
urlcolor: blue
knit: pagedown::chrome_print
---

```{css, echo = FALSE}
.remark-slide-content {
  font-size: 18px;
  padding: 20px 80px 20px 80px;
}
.remark-code, .remark-inline-code {
  background: #f0f0f0;
}
.remark-code {
  font-size: 14px;
}
.huge .remark-code { /*Change made here*/
  font-size: 200% !important;
}
.very-large .remark-code { /*Change made here*/
  font-size: 150% !important;
}
.large .remark-code { /*Change made here*/
  font-size: 125% !important;
}
.small .remark-code { /*Change made here*/
  font-size: 75% !important;
}

.very-small .remark-code { /*Change made here*/
  font-size: 50% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 40% !important;
}
```


```{r setup, include = FALSE, cache = FALSE, message = FALSE}
## set some knitr options
library(knitr)
library(here)
# ## Figure settings, defaults to png files
opts_chunk$set(fig.align='center', fig.width = 16/4,
               fig.height = 9/4, fig.show = 'hold', par = TRUE,
               dev='png',
               dpi=72*4, out.width = "80%")

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s'>%s</span>", color,
      x)
  } else x
}

hl <- function(x) {
  colorize(x, color = "corals2")
}
```

# Readings

* R for data science 
    * Introduction
    * Chapters 1 (Data visualization with `ggplot2`) and 5 (Exploratory data analysis)

---

# `data.frame` The first class object

* A `data.frame` is a rectangular collection of variables (columns) and observations (rows). 

* Built in datasets can be explored with `data()`

```{r, eval = FALSE}
data()
```

---

# Example: Palmer Penguins

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
```

.small[
```{r}
penguins
```
]

---

# First ggplot

* start by exploring the data. Use the `str()` or `glimpse()` functions to explore the data



.small[
```{r}
str(penguins)
```
]

.small[
```{r}
glimpse(penguins)
```
]

---

# flipper length and body mass

* Describe what we see

```{r ggplot}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm))
```

---

# Plotting aesthetics

* ggplot is a lot of code for a simple plot!


--

```{r base}
plot(flipper_length_mm ~ body_mass_g, data = penguins)
```

* Wasn't that easier?

---

# Aesthetics

- What if I want to color my plot by sex? 

--

- What if I want to make my circles to have size relative to bill depth?

--

- What if...

--

* Use aesthetics.

    * Aesthetics map a characteristic given a variable.
    
    * Great for quick visually communication.

---

# Using aesthetics

```{r aes-1}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm,
                             color = sex))
```


---

# Using aesthetics

```{r, aes-2}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm,
                             color = sex, size = bill_length_mm))
```

---

# Available aesthetics


* size

* color

* fill

* shape

* transparency (`alpha`)

---

# Facets

* What if I want to show multiple plots where each plot is determined by a variable?

--

* Facets.

    * `facet_wrap()` for a single variable.
    
    * `facet_grid()` for a grid of two variables.
    
---

```{r facet-1}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
    facet_wrap(~ sex)
```

---

```{r facet-2}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
    facet_grid(island ~ sex)
```


---

# Geometries

* What kind of plot do you want?

    * histograms --`geom_histogram()`
    * scatterplot -- `geom_point()`
    * boxplots -- `geom_box()`
    * dotplot -- `geom_dotplot()`
    * line plots -- `geom_line()`
    
* Each `geom` has specific aesthetic requirements

* Use the help for specifics
    * `?geom_scatter`

* Many, many others -- virtually any kind of plot you want.
    * [ggplot cheatsheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

* Many custom packages for specific plot types

---

```{r geom-1}
ggplot(data = penguins) +
    geom_histogram(mapping = aes(x = body_mass_g))
```

---

```{r geom-2}
## what happens if you use color = sex?
ggplot(data = penguins) +
    geom_histogram(mapping = aes(x = body_mass_g, fill = sex)) 
```

---
```{r geom-3}
## smooth over the density
ggplot(data = penguins) +
    geom_violin(mapping = aes(y = body_mass_g, x = sex)) 
```

---

```{r geom-4}
## what happens if you use color = sex?
ggplot(data = penguins) +
    geom_boxplot(mapping = aes(x = body_mass_g, fill = sex)) 
```

---

* You can even use multiple aesthetics

```{r geom-5}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
    geom_smooth(mapping = aes(x = body_mass_g, y = flipper_length_mm))
```

---

* geoms can have different aesthetics `aes()`.

```{r geom-6}
ggplot(data = penguins) +
    geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm,
                             color = sex)) +
    geom_smooth(mapping = aes(x = body_mass_g, y = flipper_length_mm, 
                              fill = sex, color = sex))
```

---

* Common aesthetics `aes()` can be added to the ggplot function.

```{r geom-7}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm, color = sex)) +
    geom_point() +
    geom_smooth(mapping = aes(fill = sex))
```


---

# Statistical transformations

* `ggplot` can perform statistical transformations and fit basic models

* start with `starwars` data

```{r}
data("starwars")
glimpse(starwars)
```

---

# Bar plot

```{r, bar}
ggplot(data = starwars, mapping = aes(x = sex)) +
  geom_bar()
```


---

# Bar plot using statistics

* count the number in each category of sex

```{r, count}
ggplot(data = starwars, mapping = aes(x = sex)) +
  stat_count()
```

---

# Bar plot using statistics

* count the number in each category of sex

```{r, stat-count}
ggplot(data = starwars, mapping = aes(x = sex)) +
  geom_bar(stat = "count")
```

---

# Bar plot using statistics

* The relative number in each category of sex

```{r, prop}
ggplot(data = starwars, mapping = aes(x = sex, y = stat(prop), group = 1)) +
  geom_bar()
```

---

# `stat_summary`

* plot the extent of the

```{r, stat-summary}
ggplot(data = starwars, aes(x = sex, y = height)) +
  stat_summary(
    fun.min = min,
    fun.max = max,
    fun     = mean
  )
```

---

# `stat_summary`

* Plot the quantiles

```{r, stat-summary-quantile}
ggplot(data = starwars, aes(x = sex, y = height)) +
  stat_summary(
    fun.min = function(z) quantile(z, prob = 0.1),
    fun.max = function(z) quantile(z, prob = 0.9),
    fun     = median
  )
```


---

# Plotting big data

* Use binning

```{r geom-8}
ggplot(data = penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
    geom_hex()
```

---

# Positions

* can use fill and color to highlight subsets of data

* positions of "dodge" and "jitter" can improve visualization.


.pull-left[
```{r, bar-left}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```
]

.pull-right[
```{r, bar-right}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar()
```
]


---

* position = "fill"

.pull-left[
```{r, bar-left2}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar()
```
]

.pull-right[
```{r, bar-right2}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```
]

---

* position = "identity"

    * notice these are overlapping

.pull-left[
```{r, bar-left3}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar()
```
]

.pull-right[
```{r, bar-right3}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "identity", alpha = 0.5)
```
]

---

* position = "dodge"

.pull-left[
```{r, bar-left4}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar()
```
]

.pull-right[
```{r, bar-right4}
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "dodge")
```
]

---

* position = "jitter"

    * use the built-in `mpg` dataset

```{r}
glimpse(mpg)
```


---

# Positions

* position = "jitter"

    * Useful when observations overlap

.pull-left[
```{r, bar-left5}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```
]

.pull-right[
```{r, bar-right5}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(position = "jitter")
```
]

---

# Positions

* position = "jitter"

.pull-left[
```{r, box-jitter1}
ggplot(mpg, aes(x = trans, y = hwy)) +
  geom_boxplot() +
  geom_point()
```
]

.pull-right[
```{r, box-jitter2}
ggplot(mpg, aes(x = trans, y = hwy)) +
  geom_boxplot() +
  geom_point(position = "jitter")
```
]

---

# Exploratory Data Analysis

* EDA is an artform

* Explore and learn from the data

* Guides model development

* Identifies transformations of the data that might be helpful

* Helps in formulating questions about the data

* Express creativity!

---

# Tidy data

* Each row is an observation

    * An observation is a set of measurements about an element
    
    * The element is the object on which measurement is made

* Each column is a variable

    * A variable is a characteristic of the element that can take on an value
    
* Tidy data is tabular where each row is an observation and each column is a variable

```{r}
str(diamonds)
```

# EDA

* Measures of central tendency

    * mean, median, mode

* Measures of variability

    * variance/standard deviation, range, IQR/quantiles
    
* Missing values? 

    * What do you do with missing values?
    
    * What about outliers/unusual observations?

* Covariance

* Outliers/unusal values


---

# Distributions

* Visualize marginal (one variable) distributions

    * histograms `geom_histogram()`
    
    * density plots `geom_dens()`
    
    * boxplots `geom_boxplot()`
    
    * violin plots `geom_violin()`
    

---

# Distributions
    
.pull-left[
```{r, message = FALSE, warning = FALSE}
ggplot(diamonds, aes(x = price)) +
  geom_histogram()
```
```{r}
ggplot(diamonds, aes(x = price)) +
  geom_density()
```

]


.pull-right[
```{r}
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```
```{r}
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_violin()
```
]

---

# Missing values

* As a rule, don't just ignore missing values blindly

    * I conduct a survey about income and leisure time. Why shouldn't I ignore those people that don't respond?
    
* Many functions have a `na.rm` option

```{r}
mean(penguins$bill_length_mm)
mean(penguins$bill_length_mm, na.rm = TRUE)
```
   
---

# Missing values



.pull-left[ 

* Notice the warning

```{r}
ggplot(penguins, aes(y = bill_length_mm)) +
  geom_boxplot()
```
]

.pull-right[

* No warning with `na.rm = TRUE`

```{r}
ggplot(penguins, aes(y = bill_length_mm)) +
  geom_boxplot(na.rm = TRUE)
```
]

---

# Missing values

* What if missing values are important?

* `airquality` dataset is missing the Ozone variable in June. 

.pull-left[

* Ozone measurements vs. Month

```{r warning = FALSE}
airquality %>%
  ggplot(aes(x = factor(Month), y = Ozone)) +
  geom_boxplot()
```
]

.pull-right[

* Missing values vs. Month

```{r}
airquality %>%
  mutate(missing_ozone = is.na(Ozone)) %>%
  ggplot(aes(fill = missing_ozone, x = Month)) +
  geom_density(alpha = 0.5)
```
]




---

# Covariation

* Statistical modeling is about finding patterns and covariation in data

* Is there a relationship between bill depth and bill length?

```{r, out.width = "70%", message = FALSE, warning = FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() +
  stat_smooth(method = "lm")
```

---

# Covariation

* Statistical modeling is about finding patterns and covariation in data

* Is there a relationship between bill depth and bill length?

```{r, out.width = "70%", message = FALSE, warning = FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = sex)) +
  geom_point() +
  stat_smooth(method = "lm")
```

---

# Covariation

* Statistical modeling is about finding patterns and covariation in data

* Is there a relationship between bill depth and bill length?

```{r, out.width = "70%", message = FALSE, warning = FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  stat_smooth(method = "lm")
```

---

# Covariation

* Statistical modeling is about finding patterns and covariation in data

* Is there a relationship between bill depth and bill length?

```{r, out.width = "70%", message = FALSE, warning = FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  facet_wrap(~ sex) +
  stat_smooth(method = "lm")
```

---

# Pairs plots


```{r, out.width = "70%", message = FALSE, warning = FALSE, fig.width=16/2, fig.height = 9/2}
penguins %>%
  select(species, body_mass_g, ends_with("_mm")) %>%
  GGally::ggpairs(aes(color = species)) +
  scale_colour_manual(values = c("darkorange", "purple", "cyan4")) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"))
```

---

# Namespaces

* What's the deal with the `GGally::ggpairs()` function on the last slide?

    * Often in programming there are different packages that have functions with the same name 
    * The **namespace** resolves this issue
    
    * Use the `ggpairs` function from the `GGally` package